\LaTeXe\ guide for the Journal of Functional Programming

### 2019-04-27 papers/jfp/1-intro.tex
Polishing S1-S3

In this paper we give a general framework (refined from \cite{mokhov2018buildsystems}) in which to understand and compare
build systems, in a way that is both abstract (omitting incidental detail)
and yet precise (implemented as Haskell code). With this framework we can explain confusing details from the original Shake paper \cite{mitchell2012shake} and build new features useful for the GHC build system \cite{hadrian}. Specifically we make these contributions:
\begin{itemize}
\item Build systems vary on many axes, including: static vs dynamic
  dependencies; local vs cloud; deterministic vs non-deterministic build tasks;
  support for early cutoff; self-tracking build systems; and the type of
  persistent build information. In~\S\ref{sec-background} we identify some key
  properties, illustrated by four carefully-chosen build systems.

@@

\item We show that we can instantiate our abstractions to describe the essence
of a variety of different real-life build systems, including \Make, \Shake,
\Bazel, \CloudBuild, \Buck, \Nix, and \Excel\footnote{\Excel appears very
different to the others but, seen through the lens of this paper, it is very
close indeed.}, each by the composition of the two design choices
(\S\ref{sec-implementations}). Doing this modelling in a single setting
allows the differences and similarities between these huge systems to be
brought out clearly\footnote{All our models are executable and are available on
Hackage: \url{https://hackage.haskell.org/package/build-1.0}.}.

\item Moreover, we can readily remix the ingredients to design new build systems
with desired properties, for example, to combine the advantages of \Shake and
\Bazel (\S\ref{sec-cloud-shake}), and evaluate the result (\S\ref{sec-experience}).

\end{itemize}
In short, instead of seeing build systems as unrelated
points in space, we now see them as locations in a connected landscape,
leading to a better understanding of what they do and how they compare,
and making it easier to explore other points in the
landscape.
We discuss engineering aspects in \S\ref{sec-engineering}, and related
work in \S\ref{sec-related}.

>>>>

In this paper we give a general framework in which to understand and compare
build systems, in a way that is both abstract (omitting incidental detail) and
yet precise (implemented as Haskell code). Specifically we make these
contributions:

\begin{itemize}
\item Build systems vary on many axes, including: static vs dynamic
  dependencies; local vs cloud; deterministic vs non-deterministic build tasks;
  support for early cutoff; self-tracking build systems; and the type of
  persistent build information. In~\S\ref{sec-background} we identify some key
  properties, illustrated by four carefully-chosen build systems.

@@

\item We show that we can instantiate our abstractions to describe the essence
of a variety of different real-life build systems, including \Make, \Shake,
\Bazel, \Buck, \Nix, and \Excel\footnote{\Excel appears very different to the
others but, seen through the lens of this paper, it is very close.}, each by the
composition of the two design choices~(\S\ref{sec-implementations}). Doing this
modelling in a single setting allows the differences and similarities between
these huge systems to be brought out clearly\footnote{All our models are
executable and are available on Hackage as \cmd{build-1.0}.}. We use the
abstractions to explain confusing details from the original \Shake
paper~\cite{mitchell2012shake} and build new features useful for the GHC build
system~\cite{hadrian}.

\item Moreover, we can readily remix the ingredients to design new build systems
with desired properties, for example, to combine the advantages of \Shake and
\Bazel, and evaluate the result~(\S\ref{sec-experience}).

\end{itemize}

In short, instead of seeing build systems as unrelated points in space, we now
see them as locations in a connected landscape, leading to a better
understanding of what they do and how they compare, and making it easier to
explore other points in the landscape.

We discuss engineering aspects in \S\ref{sec-engineering}, and related
work in \S\ref{sec-related}. This paper is an extended version of an earlier
conference paper~\cite{mokhov2018buildsystems}.

### 2019-04-27 papers/jfp/2-background.tex
Polishing S1-S3

\Make\footnote{There are numerous implementations of \Make and none comes with a
formal specification. In this paper we therefore use a simple and sensible
approximation to a real \Make that you might find on your machine.} was developed
more than 40 years ago to automatically build software libraries and executable
programs from source code. It uses \emph{makefiles} to describe tasks (often
referred to as \emph{build rules}) and their dependencies in a simple textual form.
For example:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{makefile}
util.o: util.h util.c
    gcc -c util.c
\end{minted}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{makefile}
main.o: util.h main.c
    gcc -c main.c
\end{minted}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{makefile}
main.exe: util.o main.o
    gcc util.o main.o -o main.exe
\end{minted}
\vspace{1mm}

\noindent
The above makefile lists three tasks: (i) compile a utility library comprising
files \cmd{util.h} and \cmd{util.c} into \cmd{util.o} by
executing\footnote{In this example we pretend \cmd{gcc} is a pure function for
the sake of simplicity. In reality, there are multiple versions of \cmd{gcc};
the actual binary used to compile and link files, and standard libraries
such that \cmd{stdio.h}, are often also listed as dependencies.} the command
\cmd{gcc -c util.c}, (ii) compile the main source file \cmd{main.c} into
\cmd{main.o}, and (iii) link object files \cmd{util.o} and \cmd{main.o} into the
executable \cmd{main.exe}. The makefile contains the complete information about
the \emph{task dependency graph}, which is shown in Fig.~\ref{fig-make}(a).

\begin{figure}%[h]
\begin{subfigure}[b]{0.32\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/make-example.pdf}}
\caption{Task dependency graph}
\end{subfigure}
\begin{subfigure}[b]{0.32\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/make-example-full.pdf}}
\caption{Full rebuild}
\end{subfigure}
\begin{subfigure}[b]{0.32\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/make-example-partial.pdf}}
\caption{Partial rebuild}
\end{subfigure}
% \vspace{-2mm}
\caption{A task dependency graph and two build scenarios. Input files are shown
in rectangles, intermediate and output files are shown in rounded rectangles.
Modified inputs and files that are rebuilt are highlighted.
\label{fig-make}}
% \vspace{-2mm}
\end{figure}

@@

\textbf{Definition} (Minimality). A build system is \emph{minimal} if it
executes tasks at most once per build and only if they transitively depend on
inputs that changed since the previous build.\label{def-minimal}
\vspace{2mm}

@@

\noindent
There are two input cells \cmd{A1} and \cmd{A2}, and a single task that computes
the sum of their values, producing the result in cell \cmd{B1}. If either of
the inputs change, \Excel will recompute the result.

Unlike \Make, \Excel does not need to know all task dependencies upfront. Indeed, some
dependencies may change \emph{dynamically} according to computation results. For
example:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{text}
A1: 10      B1: INDIRECT("A" & C1)      C1: 1
A2: 20
\end{minted}
\vspace{1mm}

\noindent
The formula in \cmd{B1} uses the \cmd{INDIRECT} function, which takes a string
and returns the value of the cell with that name.  The string evaluates to
\cmd{"A1"}, so \cmd{B1} evaluates to \cmd{10}. However, the dependencies of the
formula in \cmd{B1} are determined by the value of \cmd{C1}, so it is impossible
to compute the dependency graph before the build starts\footnote{In
this particular example one might say that the value of
\cmd{C1} is statically known, but imagine that it is the result of a long
computation chain -- its value will only become available during the build.}.

To support dynamic dependencies, \Excel's calc engine~\cite{excel_recalc} is
significantly different from \Make. \Excel arranges the cells into a linear
sequence, called the \emph{calc chain}.  During the build, \Excel processes
cells in the calc-chain sequence, but if computing a cell \cmd{C} requires the
value of a cell \cmd{D} that has not yet been computed, \Excel \emph{aborts}
computation of \cmd{C}, moves \cmd{D} before \cmd{C} in the calc chain, and
resumes the build starting with \cmd{D}. When a build is complete, the resulting
calc chain respects all the dynamic dependencies of the spreadsheet. When an
input value (or formula) is changed, \Excel uses the final calc chain from the
\emph{previous} build as its starting point so that, in the common case where
changing an input value does not change dependencies, there are no aborts.
Notice that build always succeeds regardless of the initial calc chain (barring
truly circular dependencies); the calc chain is just an optimisation.
We refer to this algorithm as \emph{restarting}, and discuss it
in more detail in~\S\ref{sec-implementation-excel}.

@@

\begin{figure}
\begin{subfigure}[b]{0.90\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/shake-example.pdf}}
\caption{Dependency graph produced after the previous build.}
\end{subfigure}
\begin{subfigure}[b]{0.90\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/shake-example-rebuild.pdf}}
\caption{Since \cmd{docs.txt} was modified, we rebuild \cmd{release.txt} and
\cmd{release.tar}, discovering a new dependency.}
\end{subfigure}
% \vspace{-2mm}
\caption{Dynamic dependencies example: create \cmd{README} and add it to the
list of release documents \cmd{docs.txt}.\label{fig-shake}}
% \vspace{-4mm}
\end{figure}

\subsection{\Shake: Dynamic Dependencies with No Remorse}
\label{sec-background-shake}

@@

\begin{itemize}
    \item \cmd{LICENSE} is an input text file containing the project license.
    \item \cmd{release.txt} is a text file listing all files that should be in the release. This file
    is produced by concatenating input files \cmd{bins.txt} and \cmd{docs.txt},
    which list all binary and documentation files of the project.
    \item \cmd{release.tar} is the release archive built by executing the
    command \cmd{tar} on the release files.
\end{itemize}

@@

\Shake's implementation is different from both \Make and \Excel in two aspects.
First, it uses the dependency graph from the previous build to decide which
files need to be rebuilt. This idea has a long history, going back to
\emph{incremental}~\cite{demers1981incremental},
\emph{adaptive}~\cite{acar2002adaptive}, and
\emph{self-adjusting computations} (see \cite{acar2007selfadjusting} and
\S\ref{sec-related}). Second, instead of aborting and deferring the execution of
tasks whose newly discovered dependencies have not yet been built (as \Excel
does), \Shake \emph{suspends} their execution until the dependencies are brought
up to date. We refer to this task scheduling algorithm as \emph{suspending},
see a concrete implementation in~\S\ref{sec-implementation-shake}.

@@

\Shake also supports the \emph{early cutoff optimisation}. When it
executes a task and the result is unchanged from the previous build, it is
unnecessary to execute the dependent tasks, and hence \Shake can stop a build
earlier, as illustrated in Fig.~\ref{fig-cutoff}. Not all build systems support
early cutoff: \Make and \Excel do not, whereas \Shake and \Bazel (introduced
below) do.

@@

\begin{figure}
% \vspace{-2mm}
\begin{subfigure}[b]{0.25\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-checkout.pdf}}
% \vspace{-1mm}
\caption{Download sources}
\end{subfigure}
\begin{subfigure}[b]{0.40\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-build.pdf}}
% \vspace{-1mm}
\caption{Build \cmd{main.exe}}
\end{subfigure}
\begin{subfigure}[b]{0.31\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-rebuild.pdf}}
% \vspace{-1mm}
\caption{Modify \cmd{util.c} and rebuild}
\end{subfigure}
% \vspace{-2.5mm}
\caption{A cloud build example: (a)~download sources, (b)~build \cmd{main.exe}
by downloading it from the cloud and skipping intermediate files (only their
hashes are needed), (c)~modify \cmd{util.c} and rebuild \cmd{main.exe}, which
requires building \cmd{util.o} (nobody has compiled \cmd{util.c} before) and
downloading \cmd{main.o} (it is needed for linking \cmd{main.exe}). File hashes
are shown in circles, and non-materialised intermediates in dashed rounded
rectangles.\label{fig-bazel}}
% \vspace{-3mm}
\end{figure}

\begin{table}[h]
% \vspace{-2mm}
\caption{Build system differences.\label{tab-summary}}
% \vspace{-4mm}
% \smaller
\centering
\begin{tabular}{llllccc}
\hline
Build system\gap & Persistent build information\gap & Scheduler\gap   & Dependencies\gap    & Minimal\gap & Cutoff\gap & Cloud\\\hline
\Make       & File modification times      & Topological\gap & Static          & Yes     & No     & No   \\
\Excel      & Dirty cells, calc chain      & Restarting  & Dynamic         & No      & No     & No   \\
\Shake      & Previous dependency graph    & Suspending  & Dynamic         & Yes     & Yes    & No   \\
\Bazel      & Cloud cache, command history\gap & Restarting  & Dynamic$^{(*)}$ & No      & Yes    & Yes  \\\hline
\end{tabular}
% \vspace{2mm}
$^{(*)}$At present, user-defined build rules cannot have dynamic dependencies.
% \vspace{-8mm}
\end{table}

% \vspace{-1.5mm}
\subsection{Summary}\label{sec-background-summary}
% \vspace{-0.5mm}

We summarise differences between four discussed build systems in
Table~\ref{tab-summary}. The column \emph{`persistent build information'} refers
to the information that build systems persistently store between builds:
\begin{itemize}
    \item \Make stores file modification times, or rather, it relies on the file
    system to do that.
    \item \Excel stores one dirty bit per cell and the calc chain from the
    previous build.
    \item \Shake stores the dependency graph discovered in the previous build,
    annotated with file content hashes for efficient checking of file changes.
    \item \Bazel stores the content-addressable cache and the history of all
    previous build commands annotated with file hashes. This information is
    shared among all users of the build system.
\end{itemize}

@@

>>>>

\Make\footnote{There are numerous implementations of \Make and none comes with a
formal specification. In this paper we use a simple approximation to a real
\Make that you might find on your machine.} was developed
more than 40 years ago to automatically build software libraries and executable
programs from source code. It uses \emph{makefiles} to describe tasks, often
referred to as \emph{build rules}, and their dependencies in a simple textual
form. For example:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{makefile}
util.o: util.h util.c
    gcc -c util.c
\end{minted}
% \vspace{1mm}
\begin{minted}[xleftmargin=10pt]{makefile}
main.o: util.h main.c
    gcc -c main.c
\end{minted}
% \vspace{1mm}
\begin{minted}[xleftmargin=10pt]{makefile}
main.exe: util.o main.o
    gcc util.o main.o -o main.exe
\end{minted}
\vspace{1mm}

\noindent
The above makefile lists three tasks: (i) compile a utility library comprising
files \cmd{util.h} and \cmd{util.c} into \cmd{util.o} by
executing\footnote{In this example we pretend \cmd{gcc} is a pure function for
the sake of simplicity. In reality, there are multiple versions of \cmd{gcc} and
the actual binary used to compile and link files, as well as standard libraries
such that \cmd{stdio.h}, are often also listed as dependencies.} the command
\cmd{gcc -c util.c}, (ii) compile the main source file \cmd{main.c} into
\cmd{main.o}, and (iii) link object files \cmd{util.o} and \cmd{main.o} into the
executable \cmd{main.exe}. The makefile contains the complete information about
the \emph{task dependency graph}, which is shown in Fig.~\ref{fig-make}(a).

\begin{figure}%[h]
\begin{subfigure}[b]{0.32\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/make-example.pdf}}
\caption{Task dependency graph}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.32\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/make-example-full.pdf}}
\caption{Full rebuild}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.32\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/make-example-partial.pdf}}
\caption{Partial rebuild}
\end{subfigure}
% \vspace{-2mm}
\caption{A task dependency graph and two build scenarios. Input files are shown
in rectangles, intermediate and output files are shown in rounded rectangles.
Modified inputs and files that are rebuilt are highlighted.
\label{fig-make}}
% \vspace{-2mm}
\end{figure}

@@

\vspace{2mm}
\textbf{Definition} (Minimality). A build system is \emph{minimal} if it
executes tasks at most once per build and only if they transitively depend on
inputs that changed since the previous build.\label{def-minimal}
\vspace{2mm}

@@

\noindent
There are two input cells \cmd{A1} and \cmd{A2}, and a single task that computes
the sum of their values, producing the result in cell \cmd{B1}. If either of
the inputs change, \Excel will recompute \cmd{B1}.

Unlike \Make, \Excel does not need to know all task dependencies upfront.
Indeed, some dependencies may change \emph{dynamically} during computation. For
example:

\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{text}
A1: 10      B1: INDIRECT("A" & C1)      C1: 1
A2: 20
\end{minted}
\vspace{0.5mm}

\noindent
The formula in \cmd{B1} uses the \cmd{INDIRECT} function, which takes a string
and returns the value of the cell with that name.  The string evaluates to
\cmd{"A1"}, so \cmd{B1} evaluates to \cmd{10}. However, the dependencies of the
formula in \cmd{B1} are determined by the value of \cmd{C1}, so it is impossible
to compute the dependency graph before the build starts. In this particular
example one might say that the value of \cmd{C1} is statically known, but
imagine that it is the result of a long computation chain -- its value will only
become available during the build.

To support dynamic dependencies, \Excel's calculation engine~\cite{excel_recalc}
is significantly different from \Make. \Excel arranges the cells into a linear
sequence, called the \emph{calc chain}.  During the build, \Excel processes
cells in the calc-chain sequence, but if computing a cell \cmd{C} requires the
value of a cell \cmd{D} that has not yet been computed, \Excel \emph{aborts}
computation of \cmd{C}, moves \cmd{D} before \cmd{C} in the calc chain, and
resumes the build starting with \cmd{D}. When a build is complete, the resulting
calc chain respects all the dynamic dependencies of the spreadsheet. When an
input value or formula is changed, \Excel uses the final calc chain from the
\emph{previous} build as its starting point so that, in the common case where
changing an input value does not change dependencies, there are no aborts.
Notice that build always succeeds regardless of the initial calc chain (barring
truly circular dependencies); the calc chain is just an optimisation.
We refer to this algorithm as \emph{restarting}, and discuss it
in more detail in~\S\ref{sec-implementation-excel}.

@@

\begin{figure}
\begin{subfigure}[b]{0.90\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/shake-example.pdf}}
\caption{Dependency graph produced after the previous build.}
\end{subfigure}
\begin{subfigure}[b]{0.90\linewidth}
\vspace{1mm}
\centerline{\includegraphics[scale=0.28]{fig/shake-example-rebuild.pdf}}
\caption{The input file \cmd{docs.txt} was modified, hence we rebuild
\cmd{release.txt} and \cmd{release.tar}, discovering a new dependency
\cmd{README} in the process.}
\end{subfigure}
% \vspace{-2mm}
\caption{Dynamic dependencies example: create \cmd{README} and add it to the
list of release documents \cmd{docs.txt}.\label{fig-shake}}
% \vspace{-4mm}
\end{figure}

\vspace{-1mm}
\subsection{\Shake: Dynamic Dependencies with No Remorse}
\label{sec-background-shake}

@@

\begin{itemize}
    \item \cmd{LICENSE} is an input text file containing the project license.
    \item \cmd{release.txt} lists all release files. This file is produced by
    concatenating input text files \cmd{bins.txt} and \cmd{docs.txt}, which list
    binary and documentation files of the project.
    \item \cmd{release.tar} is the archive built by executing the command
    \cmd{tar} on the release files.
\end{itemize}

@@

\Shake's implementation is different from both \Make and \Excel in two aspects.
First, it uses the dependency graph from the previous build to decide which
files need to be rebuilt. This idea has a long history, going back to
\emph{incremental}~\cite{demers1981incremental},
\emph{adaptive}~\cite{acar2002adaptive}, and
\emph{self-adjusting computations} -- see
Acar~\etal~\shortcite{acar2007selfadjusting} and~\S\ref{sec-related}. Second,
instead of aborting and deferring the execution of tasks whose newly discovered
dependencies have not yet been built (as \Excel does), \Shake \emph{suspends}
their execution until the dependencies are brought up to date. We refer to this
task scheduling algorithm as \emph{suspending}, see a concrete implementation
in~\S\ref{sec-implementation-shake}.

@@

\Shake also supports the \emph{early cutoff optimisation}. When it
executes a task and the result is unchanged from the previous build, it is
unnecessary to execute the dependent tasks, and hence \Shake can stop a build
earlier, as illustrated in Fig.~\ref{fig-cutoff}. Not all build systems support
early cutoff: \Shake and \Bazel (introduced below) do, but \Make and \Excel do
not.

@@

\begin{figure}
% \vspace{-2mm}
\begin{subfigure}[b]{0.25\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-checkout.pdf}}
\vspace{-0.5mm}
\caption{Download sources}
\end{subfigure}
\begin{subfigure}[b]{0.40\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-build.pdf}}
\vspace{-0.5mm}
\caption{Build \cmd{main.exe}}
\end{subfigure}
\begin{subfigure}[b]{0.31\linewidth}
\centerline{\includegraphics[scale=0.28]{fig/bazel-example-rebuild.pdf}}
\vspace{-0.5mm}
\caption{Modify \cmd{util.c} and rebuild}
\end{subfigure}
\vspace{-1mm}
\caption{A cloud build example: (a)~download sources, (b)~build \cmd{main.exe}
by downloading it from the cloud and skipping intermediate files (only their
hashes are needed), (c)~modify \cmd{util.c} and rebuild \cmd{main.exe}, which
requires building \cmd{util.o} (nobody has compiled \cmd{util.c} before) and
downloading \cmd{main.o} (it is needed for linking \cmd{main.exe}). File hashes
are shown in circles, and non-materialised intermediates in dashed rounded
rectangles.\label{fig-bazel}}
\vspace{-1mm}
\end{figure}

\vspace{-1.5mm}
\subsection{Summary}\label{sec-background-summary}
\vspace{-0.5mm}

We summarise differences between four discussed build systems in
Table~\ref{tab-summary}. The column \emph{`persistent build information'} refers
to the information that build systems persistently store between builds:
\begin{itemize}
    \item \Make stores file modification times, or rather, it relies on the file
    system to do that.
    \item \Excel stores one dirty bit per cell and the calc chain from the
    previous build.
    \item \Shake stores the dependency graph discovered in the previous build,
    annotated with file content hashes for efficient checking of file changes.
    \item \Bazel stores the content-addressable cache and the history of all
    previous build commands annotated with file hashes. This information is
    shared among all users.
\end{itemize}

@@

\begin{table}
% \vspace{-2mm}
\caption{Build system differences.\label{tab-summary}}
% \vspace{-4mm}
% \smaller
\centering
\begin{tabular}{llllccc}
\hline
Build system\gap & Persistent build information\gap & Scheduler\gap   & Dependencies\gap    & Minimal\gap & Cutoff\gap & Cloud\\\hline
\Make       & File modification times      & Topological\gap & Static          & Yes     & No     & No   \\
\Excel      & Dirty cells, calc chain      & Restarting  & Dynamic         & No      & No     & No   \\
\Shake      & Previous dependency graph    & Suspending  & Dynamic         & Yes     & Yes    & No   \\
\Bazel      & Cloud cache, command history\gap & Restarting  & Dynamic$^{(*)}$ & No      & Yes    & Yes  \\\hline
\end{tabular}
% \vspace{2mm}
$^{(*)}$At present, user-defined build rules cannot have dynamic dependencies.
% \vspace{-8mm}
\end{table}
